Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-08-25 paper/paper.tex
Continue

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=fixed,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother

@@

@@

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.
We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the type of a total function from \lstinline!'a! to \lstinline!'b!.
The types for booleans, natural numbers, and in integers are \lstinline!bool!, \lstinline!nat!, and \lstinline!int!, respectively.
Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
Analogously, we use \lstinline!'a list! to describe lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
For optional values, Isabelle/HOL offers the type \lstinline!option! where a term \lstinline!opt :: 'a option! is either \lstinline!None! or \lstinline!Some a! with \lstinline!a :: 'a!.
Finally, we remark that \textbf{iff} is equivalent to \lstinline!$=$! on type \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
\todo[inline]{Add arrows and bracket notation for Pure formulae}
\section{Syntax and Semantics of MLSS\label{sec:semantics}}
\subsection{Syntax}
At the heart of \textbf{MLSS}, we have the type of set terms which includes the empty set and variables as well as the operations union, intersection, set difference, and \lstinline!Single!, which represents a singleton set containing the argument.
\todo[inline]{Syntax for \lstinline!Single!?}
\begin{lstlisting}
datatype 'a pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\sqcup_\text{s}$ 'a pset_term
| 'a pset_term $\sqcap_\text{s}$ 'a pset_term
| 'a pset_term $-_\text{s}$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype 'a pset_atom =
  'a pset_term $\in_\text{s}$ 'a pset_term
| 'a pset_term $\approx_\text{s}$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\sqsubseteq_\text{s}$! and enumerate finite sets as \lstinline!$s \subseteq_\text{s} t$! is equivalent to \lstinline!s $\sqcup_\text{s}$ t $\approx_\text{s}$ t! and a finite set of elements \lstinline!{t$_1$,t$_2$,$\ldots$,t$_k$}! can be expressed by \lstinline!Single t$_1$ $\sqcup_\text{s}$ $\ldots$ $\sqcup_\text{s}$ Single t$_k$!.

@@

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in the formula.
Similarly, it defines a function \lstinline!set_pset_atom :: 'a pset_atom $\Rightarrow$ 'a set! with which we can extract all the variable names in a set atom.
We combine these functions in order to define \lstinline!vars_fm! that extracts all the variable names occuring in a set formula.
\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\begin{lstlisting}
fun subterms ::
    'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\sqcup_\text{s}$ t) =
    {s $\sqcup_\text{s}$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\sqcap_\text{s}$ t) =
    {s $\sqcap_\text{s}$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $-_\text{s}$ t) =
    {s $-_\text{s}$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\in_\text{s}$ t) =
    subterms s $\cup$ subterms t
| subterms_atom (s $\approx_\text{s}$ t) =
    subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (Atom a) = {Atom a}
| subfms (And p q) =
    {And p q} $\cup$ subfms p $\cup$ subfms q
| subfms (Or p q) =
    {Or p q} $\cup$ subfms p $\cup$ subfms q
| subfms (Neg q) = {Neg q} $\cup$ subfms q
\end{lstlisting}

@@

Equipped with the above, we define the interpretation function \lstinline!I$_\text{st}$! for set terms that interprets a set term with respect to a valuation function \lstinline!M :: 'a $\Rightarrow$ V! for variable.
\begin{lstlisting}
fun I$_\text{st}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  I$_\text{st}$ M $\emptyset$ = 0
| I$_\text{st}$ M (Var x) = M x
| I$_\text{st}$ M (Single s) = vset {I$_\text{st}$ M s}
| I$_\text{st}$ M (s $\sqcup_\text{s}$ t) = I$_\text{st}$ M s $\sqcup$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $\sqcap_\text{s}$ t) = I$_\text{st}$ M s $\sqcap$ I$_\text{st}$ M t 
| I$_\text{st}$ M (s $-_\text{s}$ t) = I$_\text{st}$ M s $-$ I$_\text{st}$ M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun I$_\text{sa}$ :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ V where
  I$_\text{sa}$ M (s $\in_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s $\in$ elts (I$_\text{st}$ M t)
| I$_\text{sa}$ M (s $\approx_\text{s}$ t) $\longleftrightarrow$ I$_\text{st}$ M s = I$_\text{st}$ M t
\end{lstlisting}
For formulae, we write \lstinline!M $\models$ $\phi$! for the judgement that $\phi$ holds under the valuation function \lstinline!M!.
We omit the implementation of $\models$ for brevity.

\section{A Tableau Calculus for MLSS}
\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

@@

\begin{lstlisting}
definition params' :: 'a branch $\Rightarrow$ 'a set where
  params' b $\equiv$
    {c $\in$ params b. $\forall$t $\in$ subterms_fm (last b).
      AT (Var c $\approx_\text{s}$ t) $\notin$ set b $\land$
      AT (t $\approx_\text{s}$ Var c) $\notin$ set b} 
\end{lstlisting}

>>>>

\usepackage{listings}
\definecolor{isarblue}{HTML}{006699}
\definecolor{isargreen}{HTML}{009966}
\lstdefinelanguage{isabelle}{%
    keywords=[1]{type_synonym,datatype,fun,function,abbreviation,definition,proof,lemma,theorem,corollary,inductive},
    keywordstyle=[1]\bfseries\color{isarblue},
    keywords=[2]{where,assumes,shows,and},
    keywordstyle=[2]\bfseries\color{isargreen},
    keywords=[3]{if,then,else,case,of,SOME,let,in,O},
    keywordstyle=[3]\color{isarblue},
}
\lstset{%
  language=isabelle,
  escapeinside={&}{&},
  columns=fixed,
  extendedchars,
  basewidth={0.5em,0.45em},
  basicstyle=\ttfamily,
  mathescape,
}
\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.4\baselineskip}}
\makeatother

@@

% Commands
\newlength{\trianglewidth}
\settowidth{\trianglewidth}{\(\triangleleft\)}

\newcommand{\lefttrianglebar}{\mathrel{\tikz[baseline]{\draw (1ex, 0.75ex) -- (0, 1.25ex) -- (0, 0.25ex) -- cycle; \draw (0, 0.75ex) -- (1ex, 0.75ex);}}}
\newcommand{\lefttriangle}{\mathrel{\tikz[baseline]{\draw (1ex, 0.75ex) -- (0, 1.25ex) -- (0, 0.25ex) -- cycle;}}}

\newcommand{\sextends}{\lefttriangle}
\newcommand{\fextends}{\lefttrianglebar}

\newcommand{\unionS}{\sqcup_\text{s}}
\newcommand{\interS}{\sqcap_\text{s}}
\newcommand{\diffS}{-_\text{s}}
\newcommand{\inS}{\in_\text{s}}
\newcommand{\notinS}{\notin_\text{s}}
\newcommand{\eqS}{\approx_\text{s}}
\newcommand{\neqS}{{\not\approx}_\text{s}}
\newcommand{\subseteqS}{\sqsubseteq_\text{s}}
\newcommand{\Ist}{I$_\text{st}$}
\newcommand{\Isa}{I$_\text{sa}$}

@@

\subsection{Notation}
Isabelle/HOL~\cite{isabelle} conforms to everyday mathematical notation for the most part.
For the benefit of the reader that is unfamiliar with Isabelle/HOL, we establish notation and in particular some essential datatypes together with their primitive operations that are specific to Isabelle/HOL.
We write \lstinline!t :: 'a! to specify that the term \lstinline!t! has the type \lstinline!'a! and \lstinline!'a $\Rightarrow$ 'b! for the type of a total function from \lstinline!'a! to \lstinline!'b!.
The types for booleans, natural numbers, and in integers are \lstinline!bool!, \lstinline!nat!, and \lstinline!int!, respectively.
Sets with elements of type \lstinline!'a! have the type \lstinline!'a set!.
Analogously, we use \lstinline!'a list! to describe lists, which are constructed as the empty list \lstinline![]! or with the infix constructor \lstinline!#!, and are appended with the infix operator \lstinline!@!.
The function \lstinline!set! converts a list into a set.
For optional values, Isabelle/HOL offers the type \lstinline!option! where a term \lstinline!opt :: 'a option! is either \lstinline!None! or \lstinline!Some a! with \lstinline!a :: 'a!.
Finally, we remark that \textbf{iff} is equivalent to \lstinline!$=$! on type \lstinline!bool! and \lstinline!$\equiv$! is definitional equality of the meta-logic of Isabelle/HOL, which is called Isabelle/Pure.
\todo[inline]{Add arrows and bracket notation for Pure formulae}
\section{Syntax and Semantics of MLSS\label{sec:semantics}}
\subsection{Syntax}
At the heart of \textbf{MLSS}, we have the type of set terms which includes the empty set and variables as well as the operations union, intersection, set difference, and \lstinline!Single!, which represents a singleton set containing the argument.
\todo[inline]{Syntax for \lstinline!Single!?}
\begin{lstlisting}
datatype 'a pset_term =
  $\emptyset$ | Var 'a | Single ('a pset_term)
| 'a pset_term $\unionS$ 'a pset_term
| 'a pset_term $\interS$ 'a pset_term
| 'a pset_term $\diffS$ 'a pset_term
\end{lstlisting}
We can combine two set terms to form a set atom by using the membership or the equality operator.
\begin{lstlisting}
datatype 'a pset_atom =
  'a pset_term $\inS$ 'a pset_term
| 'a pset_term $\eqS$ 'a pset_term
\end{lstlisting}
With the above operators we can also represent the subset operator \lstinline!$\subseteqS$! and enumerate finite sets as \lstinline!s $\subseteqS$ t! is equivalent to \lstinline!s $\unionS$ t $\eqS$ t! and a finite set of elements \lstinline!{t$_1$,t$_2$,$\ldots$,t$_k$}! can be expressed by \lstinline!Single t$_1$ $\unionS$ $\ldots$ $\unionS$ Single t$_k$!.

@@

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
We will sometimes drop the \lstinline!Atom! constructor to improve presentation.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!Neg (Atom (s $\inS$ t))! and \lstinline!Neg (Atom (s $\eqS$ t))!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in the formula.
Similarly, it defines a function \lstinline!set_pset_atom :: 'a pset_atom $\Rightarrow$ 'a set! with which we can extract all the variable names in a set atom.
We combine these functions in order to define \lstinline!vars_fm! that extracts all the variable names occuring in a set formula.
\begin{lstlisting}
definition vars_fm :: 'a pset_fm $\Rightarrow$ 'a set where
  vars_fm $\phi$ $\equiv$ $\bigcup$(set_pset_atom ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\begin{lstlisting}
fun subterms ::
    'a pset_term $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms_atom ::
      'a pset_atom $\Rightarrow$ 'a pset_term set where
  subterms_atom (s $\inS$ t) =
    subterms s $\cup$ subterms t
| subterms_atom (s $\eqS$ t) =
    subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms_fm ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms_fm $\phi$ $\equiv$ $\bigcup$(subterms_atom ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (Atom a) = {Atom a}
| subfms (And p q) =
    {And p q} $\cup$ subfms p $\cup$ subfms q
| subfms (Or p q) =
    {Or p q} $\cup$ subfms p $\cup$ subfms q
| subfms (Neg q) = {Neg q} $\cup$ subfms q
\end{lstlisting}

@@

Equipped with the above, we define the interpretation function \lstinline!I_$\text{st}$! for set terms that interprets a set term with respect to a valuation function \lstinline!M :: 'a $\Rightarrow$ V! for variable.
\begin{lstlisting}
fun &\Ist& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_term $\Rightarrow$ V where
  &\Ist& M $\emptyset$ = 0
| &\Ist& M (Var x) = M x
| &\Ist& M (Single s) = vset {&\Ist& M s}
| &\Ist& M (s $\unionS$ t) = &\Ist& M s $\sqcup$ &\Ist& M t 
| &\Ist& M (s $\interS$ t) = &\Ist& M s $\sqcap$ &\Ist& M t 
| &\Ist& M (s $\diffS$ t) = &\Ist& M s $-$ &\Ist& M t 
\end{lstlisting}
The interpretation function \lstinline!I$_\text{sa}$! for set atoms is straightforward as well.
\begin{lstlisting}
fun &\Isa& :: ('a $\Rightarrow$ V) $\Rightarrow$ 'a pset_atom $\Rightarrow$ bool
where
  &\Isa& M (s $\inS$ t) $\longleftrightarrow$ &\Ist& M s $\in$ elts (&\Ist& M t)
| &\Isa& M (s $\eqS$ t) $\longleftrightarrow$ &\Ist& M s = &\Ist& M t
\end{lstlisting}
For formulae, we write \lstinline!M $\models$ $\phi$! for the judgement that $\phi$ holds under the valuation function \lstinline!M!.
We omit the implementation of $\models$ for brevity.

\section{A Tableau Calculus for MLSS}
$A \fextends B \sextends C$

\begin{lstlisting}
type_synonym 'a branch = 'a pset_fm list
\end{lstlisting}

@@

\begin{lstlisting}
inductive bclosed :: 'a branch $\Rightarrow$ bool where
  $\llbracket$ $\phi$ $\in$ set b; Neg $\phi$ $\in$ set b $\rrbracket$ $\Longrightarrow$ bclosed b
| (t $\inS$ $\emptyset$) $\in$ set b $\Longrightarrow$ bclosed b
| (t $\neqS$ t) $\in$ set b $\Longrightarrow$ bclosed b
| $\llbracket$ member_cycle cs; set cs $\subseteq$ set b$\rrbracket$
  $\Longrightarrow$ bclosed b
\end{lstlisting}

@@

\begin{lstlisting}
definition params' :: 'a branch $\Rightarrow$ 'a set where
  params' b $\equiv$
    {c $\in$ params b. $\forall$t $\in$ subterms_fm (last b).
      AT (Var c $\eqS$ t) $\notin$ set b $\land$
      AT (t $\eqS$ Var c) $\notin$ set b} 
\end{lstlisting}
